A Plasma Torus Around a Young Low-Mass Star

### 2025-04-04 paper/ms.tex
update

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that 
on average are subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to large doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

@@

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  The main argument against a starspot-only
explanation invokes the timescales and amplitudes of the sharpest
photometric features.  However, no independent evidence has yet been
acquired for the presence of any circumstellar material.
Geometrically, transiting clumps would imply an occurrence a few to at
most ten times the observed rate.  The clumps could therefore exist
around 10-30\% of M dwarfs during their early lives.

@@

% photometry
Figure~\ref{fig:spec} compiles the TESS and HIRES data from February
2024.  Over timescales of years, CPVs maintain a fixed period while
their photometric morphology evolves.  TIC~141146667 indeed evolved
relative to the February 2022 discovery data (see Figure~\ref{fig:lc}).
In February 2024, the average photometric signal showed a small
brightening over 45\% of the period, followed by a complex flux dip
spanning 55\% of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in forward models of warped
plasma tori (see \cite{Townsend2008} and the associated
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}).
% particularly ω/ωc = 0.5, beta = 30, i = 90.

@@

% * keck/hires: behavior at v/veq<1
The H$\alpha$ line core is more complex.  At $|\Delta v /
v_{\rm eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material may additionally modulate the
line profile.
Figure~\ref{fig:spec}e suggests line core variability caused by both
bright and dark regions on the star's surface, superposed
with smaller-amplitude variability from the transiting circumstellar
material.
For instance, from $\phi$=0-0.3, the double-peaked emission feature
is visible when it is being viewed both on and off-limb.
However, the large emission feature that crosses the star from
$\phi$=0.4-0.9 cannot be circumstellar in origin; it emits at an
amplitude greater than that observed from the circumstellar
components, and it crosses the stellar velocity surface too slowly.
It must therefore originate from a chromospherically bright region
near the stellar surface.
While the origin of the other bright and dark streaks passing across the
line core are more ambiguous,
a final exercise to quantify the behavior of this region is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

\begin{table*}
\small
\setlength{\tabcolsep}{2pt}
\centering
\caption{Selected System Parameters for TIC~141146667}
\label{tab:params}
\begin{tabular}{llcc}
\hline \hline
Parameter & Description & Value & Source\\
\hline 
%
$T_{\rm eff}$\dotfill                   & Effective Temperature (K) \hspace{9pt}\dotfill                 & 2972 $\pm$ 40    & 1 \\
%
$R_\star$\dotfill                       & Stellar radius ($R_\odot$)\dotfill                             & 0.42$\pm$0.02    & 1 \\
%
Age                                     & Stellar age range (Myr)\dotfill                                & 35-150           & 2 \\
%
$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02    & 3 \\
%
$\gamma$\dotfill                        & Systemic radial velocity (\kms)\dotfill                        & 0.61 $\pm$ 1.47  & 4 \\
%
Spec. Type\dotfill                      & Spectral Type\dotfill                                          & M5.5Ve           & 4 \\
%
$P_{\rm rot}$\dotfill                   & Rotation period (hr)\dotfill                                   & $3.930\pm 0.001$ & 5 \\
%
$v_{\rm eq}$\dotfill		                & Equatorial velocity \dotfill                                   &  130$\pm$4       & 6 \\
                                        & \hspace{3pt} ($2\pi R_\star/P_{\rm rot}$) (\kms)	             &                      \\
%
$v_{\rm eq}\sin{i_\star}$\dotfill		    & Projected rotational\dotfill                                   &  138$\pm$8       & 4 \\
                                        & \hspace{3pt} velocity (\kms)	                                 &                      \\
%
$v_{\rm break}$\dotfill		              & Breakup velocity \dotfill                                      &  316$\pm$16      & 6 \\
                                        & \hspace{3pt} ($G M_\star / R_\star$)$^{1/2}$ (\kms)	           &                      \\
%
$i_\star$\dotfill                       & Stellar inclination\dotfill                                    & 	$>$63           & 4 \\
                                        & \hspace{3pt}  2$\sigma$ lower limit (deg)	                     &                      \\
%
$d$\dotfill                             & Distance (pc)\dotfill                                          & $57.54 \pm 0.09$ & 7 \\
%
$R_{\rm c}$\dotfill		                  & Keplerian corotation\dotfill                                   & $1.82 \pm 0.10$  & 6 \\
                                        & \hspace{3pt} radius ($R_\star$)	                               &                      \\
%
$a_1$\dotfill                           & Clump 1 orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.1-2.7         & 4 \\
$a_2$\dotfill                           & Clump 2 orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  3.9             & 4 \\
%
$\langle$EW$_{\rm H\alpha}$$\rangle$    & Time-averaged H$\alpha$ line core                              &  7.2 $\pm$ 0.2   & 4 \\ 
                                        & \hspace{3pt} equivalent width (\AA)	                           &                      \\
\hline
\end{tabular}
\begin{flushleft}
\footnotesize{ \textsc{NOTE}---
Provenances are:
1: SED fit \cite{Bouma2024}.
2: Gaia DR3 photometry shows the star is on the pre-main sequence,
   while the spectrum lacks lithium (see~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.
4: Keck/HIRES.
5: TESS light curve.
6: Derived quantity.
7: Gaia DR3 geometric \cite{GaiaDR3}.
}
\end{flushleft}
\vspace{-0.5cm}
\end{table*}

@@

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin for the CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, where $n$ denotes the principal quantum number.
More quantitatively, the observed H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see Section~\ref{subsec:gas}).  Dust, if
present, is independently constrained to have a total mass
$10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an old and undetected disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce broadband photometric variability through bound-free
scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  
The plasma accumulates at antipodes 180$^\circ$ apart because the
deepest local minima in the effective potential exist along the line
of intersection between the rotational and magnetic equators (see
Equation~22 of \cite{Townsend2005}).
For $\sigma$~Ori~E and almost
all of its analogs, the result is ``simple'' light curves that 
resemble those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma, or else a dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical 
suggest a chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such
as electron scattering in a plasma transiting over starspots might
also produce chromatic features \cite{Rackham2018}.  The composition
and size distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-20\,$\mu$m.  Composition and size distributions similar to debris
from rocky bodies seen around white dwarfs \cite{Reach2009} would
indicate a rocky-body origin.  Compositions and sizes similar to the
interstellar medium could be indicative of condensed dust in an M
dwarf wind, similar to that formed in the environments of more evolved
stars \cite{Marigo2008}.

@@

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material
transit the star.  The implied geometric correction suggests that an
appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host similar
circumstellar environments to the CPVs.
Future studies that combine spectroscopic, polarimetric, and
multi-wavelength observations, along with magnetohydrodynamic
modeling, will be key to understanding the complex magnetospheric
environments of these young stars.

@@

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours
during the second half of the night, spanning 17 February 2024 10:47
to 16:13 (UT).
The star's airmass over this window spanned $z$=1.2-2.2, and we opted
for a fixed 15 minute cadence over the entire sequence, except for a
final 10 minute exposure due to increasing sky brightness at sunrise.
We observed without the iodine cell and used the C2 decker
(0$\farcs$86$\times$14$\farcs$0) in the red instrument configuration,
yielding a spectral resolution $R$$\approx$45{,}0000 ($\delta
v$$\approx$6.7\,\kms; $\delta v / v_{\rm eq}$$\approx$0.05).  We binned
the CCD readout by a factor of three in the spatial dimension, yielding
$\approx$1,000 photons (S/N=33) per pixel in the continuum at 6500\,\AA,
at minimum airmass.  Strong winds contributed to 1\farcs2$\pm$0\farcs2
seeing over the night, but conditions were otherwise favorable.  We
reduced the echelleogram to a one-dimensional spectrum using the
standard techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

{\it Radial Velocity}---We measured radial velocities of TIC~141146667
from our HIRES spectra using a pipeline that we developed for
rapidly rotating stars.  Our method is based on template-matching
against synthetic spectra produced by the PHOENIX stellar atmosphere
code \cite{Husser2013}.  We used the PHOENIX models with solar
metallicity and alpha element abundances, and calibrated our pipeline
using the standard stars described by \cite{Chubak2012}.  We used
velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.  We used \texttt{barycorrpy}
\cite{Kanodia2018} to calculate the velocity corrections due to Earth's
motion around the solar system barycenter and due to Earth's daily
rotation about its axis.  Our analysis code reproduces the systemic
velocities of known velocity standards \cite{Chubak2012} with an RMS of
0.66\,\kms.

@@

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been studied
for over sixty years \cite{Hayashi1963,Bildsten1997,Burke2004}.
\cite{Wood2023} provided a recent overview: an abbreviated
summary is that sufficiently cool and young M dwarfs show the 6707.8\,\AA\
doublet in absorption, $\gtrsim$10\% below their ``continua''.  Unlike
for Sun-like stars, the continuum for M dwarfs is poorly defined due to
their molecular absorption.  We therefore attempted a lithium measurement by
constructing a wavelength-binned and Doppler-corrected TIC~141146667
spectrum, assigning its uncertainties based on the measured scatter
across the five hour dataset.  We then compared this average spectrum
against the nearest matching M6 template from \cite{Bochanski2007}.  The
data show a small depression near the expected lithium wavelength,
potentially consistent with the $\Delta \lambda$$\approx$3\,\AA\ line
broadening.  This feature nominally yields ${\rm EW}_{\rm
Li}$=71$^{+18}_{-13}$\,m\AA, where the statistical uncertainties are
evaluated using a bootstrap resampling technique from the statistical
uncertainties in the HIRES spectrum.  However, the systematic uncertainties associated
with the continuum normalization are likely comparable to
the amplitude of this feature; we therefore treat the result of this
measurement as a $2\sigma$ upper limit: ${\rm EW}_{\rm Li}$$<$107\,m\AA.

@@

Extended Data Figure~\ref{fig:hirescuts} shows a few regions of interest in the
HIRES spectra, which cover 3650-7960\,\AA.  Higher order Balmer lines
including H$\gamma$ and H$\delta$ ($n=5\rightarrow2$ and
$n=6\rightarrow2$) do show variability outside the line core.
However, this variability is not clearly periodic in the same manner
as the emission seen in H$\alpha$.  This could be because there are
insufficient hydrogen ions in the relevant excited states, or because
the spectra have lower precision in these bluer regions.  The Ca HK
doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are sufficiently blended to be rendered
unusable.

@@

{\it A model for the time-dynamic H$\alpha$ spectrum}---While
Figure~\ref{fig:spec} shows that clumps of circumstellar material
exist around TIC~141146667, it would be valuable to quantify the exact
orbital periods, velocities, and velocity dispersions of these clumps.
These quantities can clarify the precision with which the
spectroscopic period agrees with the photometric period, and also
constrain the physical dimensions of the emitting region.

Given that a full radiative transfer simulation was outside our scope,
we opted to construct a phenomenological model aimed at capturing 
the emission from the circumstellar material.  We did this by
fitting each spectral epoch with a multi-component
gaussian, after having subtracted the time-average line profile as in
Figure~\ref{fig:spec}e.  The results of this exercise are shown in
Figure~\ref{fig:halphamodel}; details in its implementation and
interpretation follow.

>>>>

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that 
on average are subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to high doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

@@

The two leading hypotheses for explaining CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  The main argument against a starspot-only
explanation invokes the timescales and amplitudes of the sharpest
photometric features.  However, no independent evidence has yet been
acquired for the presence of any circumstellar material.
Geometrically, transiting clumps would imply an occurrence a few to at
most ten times the observed rate.  The clumps could therefore exist
around 10-30\% of M dwarfs during their early lives.

@@

% photometry
Figure~\ref{fig:spec} compiles the TESS and HIRES data from February
2024.  Over timescales of years, CPVs maintain fixed periods while
their photometric morphology evolves.  TIC~141146667 indeed evolved
relative to the February 2022 discovery data (see Figure~\ref{fig:lc}).
In February 2024, the average photometric signal showed a small
brightening over 45\% of the period, followed by a complex flux dip
spanning 55\% of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in forward models of warped
plasma tori (see \cite{Townsend2008} and the associated
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}).
% particularly ω/ωc = 0.5, beta = 30, i = 90.

@@

% * keck/hires: behavior at v/veq<1
The H$\alpha$ line core is more complex.  At $|\Delta v /
v_{\rm eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material might also modulate the
line profile.
Figure~\ref{fig:spec}e suggests line core variability caused by both
bright and dark regions on the star's surface, superposed
with smaller-amplitude variability from the transiting circumstellar
material.
For instance, from $\phi$=0-0.3, the double-peaked emission feature
is visible when it is being viewed both on and off-limb.
However, the large emission feature that crosses the star from
$\phi$=0.4-0.9 cannot be circumstellar in origin; it emits at an
amplitude greater than that observed from the circumstellar components,
and it crosses the stellar velocity surface too slowly to be
circumstellar.
It must therefore originate from a chromospherically bright region
on the star's surface.
While the origin of the other bright and dark streaks passing across the
line core are more ambiguous,
a final exercise to quantify the behavior of this region is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

\begin{table*}
\small
\setlength{\tabcolsep}{2pt}
\centering
\caption{Selected System Parameters for TIC~141146667}
\label{tab:params}
\begin{tabular}{llcc}
\hline \hline
Parameter & Description & Value & Source\\
\hline 
%
$T_{\rm eff}$\dotfill                   & Effective Temperature (K) \hspace{9pt}\dotfill                 & 2972 $\pm$ 40    & 1 \\
%
$R_\star$\dotfill                       & Stellar radius ($R_\odot$)\dotfill                             & 0.42$\pm$0.02    & 1 \\
%
Age                                     & Stellar age range (Myr)\dotfill                                & 35-150           & 2 \\
%
$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02    & 3 \\
%
$\gamma$\dotfill                        & Systemic radial velocity (\kms)\dotfill                        & 0.61 $\pm$ 1.47  & 4 \\
%
Spec. Type\dotfill                      & Spectral Type\dotfill                                          & M5.5Ve           & 4 \\
%
$P_{\rm rot}$\dotfill                   & Photometric rotation period (hr)\dotfill                       & $3.930\pm 0.001$ & 5 \\
%
$v_{\rm eq}$\dotfill		                & Equatorial velocity \dotfill                                   &  130$\pm$4       & 6 \\
                                        & \hspace{3pt} ($2\pi R_\star/P_{\rm rot}$) (\kms)	             &                      \\
%
$v_{\rm eq}\sin{i_\star}$\dotfill		    & Projected rotational\dotfill                                   &  138$\pm$8       & 4 \\
                                        & \hspace{3pt} velocity (\kms)	                                 &                      \\
%
$v_{\rm break}$\dotfill		              & Breakup velocity \dotfill                                      &  316$\pm$16      & 6 \\
                                        & \hspace{3pt} ($G M_\star / R_\star$)$^{1/2}$ (\kms)	           &                      \\
%
$i_\star$\dotfill                       & Stellar inclination\dotfill                                    & 	$>$63           & 4 \\
                                        & \hspace{3pt}  2$\sigma$ lower limit (deg)	                     &                      \\
%
$d$\dotfill                             & Distance (pc)\dotfill                                          & $57.54 \pm 0.09$ & 7 \\
%
$R_{\rm c}$\dotfill		                  & Keplerian corotation\dotfill                                   & $1.82 \pm 0.10$  & 6 \\
                                        & \hspace{3pt} radius ($R_\star$)	                               &                      \\
%
$r_0$\dotfill                           & Inner clump (0) orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.07$\pm$0.04                 & 4 \\
$r_1$\dotfill                           & Inner clump (1) orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  2.88$\pm$0.10                 & 4 \\
$r_2$\dotfill                           & Outer clump orbital radius ($R_\star$)\hspace{9pt}\dotfill         &  4.16$\pm$0.56                 & 4 \\
%
$\langle$EW$_{\rm H\alpha}$$\rangle$    & Time-averaged H$\alpha$ line core                              &  7.2 $\pm$ 0.2   & 4 \\ 
                                        & \hspace{3pt} equivalent width (\AA)	                           &                      \\
\hline
\end{tabular}
\begin{flushleft}
\footnotesize{ \textsc{NOTE}---
Provenances are:
1: SED fit \cite{Bouma2024}.
2: Gaia DR3 photometry shows the star is on the pre-main sequence,
   while the spectrum lacks lithium (see \S~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.
4: Keck/HIRES (see \S~\ref{subsec:halpha}).
5: TESS light curve.
6: Derived quantity.
7: Gaia DR3 geometric \cite{GaiaDR3}.
}
\end{flushleft}
\vspace{-0.5cm}
\end{table*}

@@

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin for the CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, where $n$ denotes the principal quantum number.
More quantitatively, the observed H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see Section~\ref{subsec:gas}).  Dust, if
present, is independently constrained to have a total mass
$10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an undetected old disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds which
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce broadband photometric variability through bound-free
scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  
The plasma accumulates at antipodes 180$^\circ$ apart because the
deepest local minima in the effective potential exist along the line
of intersection between the rotational and magnetic equators (see
Equation~22 of \cite{Townsend2005}).
For $\sigma$~Ori~E and almost
all of its analogs, the result is ``simple'' light curves that 
resemble those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma or dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical 
suggest a chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such
as electron scattering in a plasma transiting over starspots might
also produce chromatic features \cite{Rackham2018}.  The composition
and size distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-20\,$\mu$m.  Composition and size distributions similar to debris
from rocky bodies seen around white dwarfs \cite{Reach2009} would
indicate a rocky-body origin.  Compositions and sizes similar to the
interstellar medium could be indicative of condensed dust in an M
dwarf wind, similar to that formed around more evolved stars
\cite{Marigo2008}.

@@

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material
transit the star.  The implied geometric correction suggests that an
appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host 
circumstellar environments similar to the CPVs.
Future studies that combine spectroscopic, polarimetric, and
multi-wavelength observations, along with magnetohydrodynamic
modeling, will be key to understanding the complex magnetospheric
environments of these young stars.

@@

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours
during the second half of the night, spanning 17 February 2024 10:47
to 16:13 (UT).
The star's airmass over this window spanned $z$=1.2-2.2, and we opted
for a fixed 15 minute cadence over the entire sequence, except for a
final 10 minute exposure due to increasing sky brightness at sunrise.
We observed without the iodine cell and used the C2 decker
(0$\farcs$86$\times$14$\farcs$0) in the red instrument configuration,
yielding a spectral resolution $R$$\approx$45{,}000 ($\delta
v$$\approx$6.7\,\kms; $\delta v / v_{\rm eq}$$\approx$0.05).  We binned
the CCD readout by a factor of three in the spatial dimension, yielding
$\approx$1,000 photons (S/N=33) per pixel in the continuum at 6500\,\AA,
at minimum airmass.  Strong winds contributed to 1\farcs2$\pm$0\farcs2
seeing over the night, but conditions were otherwise favorable.  We
reduced the echelleogram to a one-dimensional spectrum using the
standard techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

{\it Radial Velocity}---We measured the radial velocities of TIC~141146667
from our HIRES spectra using a pipeline that we developed for
rapidly rotating stars.  Our method is based on template-matching
against synthetic spectra produced by the PHOENIX stellar atmosphere
code \cite{Husser2013}.  We used the PHOENIX models with solar
metallicity and alpha element abundances, and calibrated our pipeline
using the standard stars described by \cite{Chubak2012}.  We used
velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.  We used \texttt{barycorrpy}
\cite{Kanodia2018} to calculate the velocity corrections due to Earth's
motion around the solar system barycenter and due to Earth's daily
rotation about its axis.  Our analysis code reproduces the systemic
velocities of known velocity standards \cite{Chubak2012} with an RMS of
0.66\,\kms.

@@

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been studied
for over sixty years \cite{Hayashi1963,Bildsten1997,Burke2004}.
\cite{Wood2023} provided a recent overview: an abbreviated
summary is that sufficiently cool and young M dwarfs show the 6707.8\,\AA\
doublet in absorption, $\gtrsim$10\% below their continua.  However unlike
for Sun-like stars, the continuum for M dwarfs is challenging to define due to
their molecular absorption.  We therefore attempted a lithium measurement by
constructing a wavelength-binned and Doppler-corrected TIC~141146667
spectrum, assigning its uncertainties based on the measured scatter
across the five hour dataset.  We then compared this average spectrum
against the nearest matching M6 template from \cite{Bochanski2007}.  The
data show a small depression near the expected lithium wavelength,
potentially consistent with the $\Delta \lambda$$\approx$3\,\AA\ line
broadening.  This feature nominally yields ${\rm EW}_{\rm
Li}$=71$^{+18}_{-13}$\,m\AA, where the statistical uncertainties are
evaluated using a bootstrap resampling technique from the statistical
uncertainties in the HIRES spectrum.  However, the systematic uncertainties associated
with the continuum normalization are likely comparable to
the amplitude of this feature; we therefore treat the result of this
measurement as a $2\sigma$ upper limit: ${\rm EW}_{\rm Li}$$<$107\,m\AA.

@@

Extended Data Figure~\ref{fig:hirescuts} shows a few regions of interest in the
HIRES spectra, which cover 3650-7960\,\AA.  Higher order Balmer lines
including H$\gamma$ and H$\delta$ ($n=5\rightarrow2$ and
$n=6\rightarrow2$) do show variability outside the line core.
However, this variability is not clearly periodic in the same manner
as the emission seen in H$\alpha$.  This could be because there are
insufficient hydrogen ions in the relevant excited states, or because
the spectra have lower precision in these bluer regions.  The Ca HK
doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are too blended to be usable.

@@

{\it A model for the time-dynamic H$\alpha$ spectrum}---While
Figure~\ref{fig:spec} shows that clumps of circumstellar material
exist around TIC~141146667, there is value in quantifying the exact
orbital periods, velocities, and velocity dispersions of these clumps.
These quantities can constrain the physical dimensions of the emitting
region, and can also clarify the degree to which the spectroscopic
period agrees with the photometric period.

Given that a full radiative transfer simulation was outside our scope,
we opted to construct a phenomenological model aimed at capturing 
the emission from the circumstellar material.  We did this by
fitting each spectral epoch with a multi-component
gaussian, after having subtracted the time-average line profile as in
Figure~\ref{fig:spec}e.  The results of this exercise are shown in
Figure~\ref{fig:halphamodel}; details in the implementation and
interpretation follow.

To implement the model, we assumed that the ``inner'' ($K_{\rm
inner}$$\approx$2.5\,$v_{\rm eq}$) clump would be well-fit by a sum of
two gaussians because it is visually double-peaked in the raw
data from $\phi$=0.15-0.35 and $\phi$=0.65-0.85
(Figure~\ref{fig:spec}b).  We assumed that the ``outer'' ($K_{\rm
outer}$$\approx$3.9\,$v_{\rm eq}$) clump would be better fit by a
single gaussian, based on its behavior from $\phi$=0.6-0.9.  Each
gaussian component has three free parameters at each spectral epoch:
the mean $\mu$, standard deviation $\sigma$, and amplitude $A$.
We opted to label the inner component's two gaussians $i=\{ 0, 1 \}$,
and the single outer component as $i=2$.  Given the complexity of the
line profile data (Figure~\ref{fig:halphamodel} left column), the
likelihood function for this model is multimodal.  We therefore
imposed the prior constraints that $A_i \sim \mathcal{U}[0,1]$,
$\sigma_i/v_{\rm eq} \sim \mathcal{U}[0,1]$, and further assumed
$\mu_i(t) \sim
\mathcal{U}[
  K_{\rm inner}\sin(\phi(t)) - v_{\rm eq},
  K_{\rm inner}\sin(\phi(t)) + v_{\rm eq}
]$
for the inner two components, and
$\mu_2(t) \sim
\mathcal{U}[
  K_{\rm outer}\sin(\phi(t) + \pi) - 2v_{\rm eq},
  K_{\rm outer}\sin(\phi(t) + \pi) + 2v_{\rm eq}
]$
for the outer component.  This prior on the means mitigates 
multimodality in the likelihood by requiring the mean velocity of each
component to be within a one or two $v_{\rm eq}$ of the time-variable
sinusoid suggested by visual inspection.  We fitted each component to
the data {\it independently} using scipy's non-linear least squares
\texttt{curve\_fit} implementation \cite{Virtanen2020}, and scaled the
resulting parameter covariance matrix by a constant factor to match
the sample variance of the residuals.  The resulting means,
amplitudes, and standard deviations for each component are given in
Extended Data Table~\ref{tab:halphamodelparams}.
